---
title: "IPUMS API"
author: "Chris Day"
date: "2024-01-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries}
library(ipumsr)
library(tidyverse)
library(sf)
library(purrr)
library(mapview)
```

## Useful Sites
 - [Time Series Explanation](https://www.nhgis.org/time-series-tables#layouts)
 - [Data Download](https://data2.nhgis.org/main)
 - [R API](https://tech.popdata.org/ipumsr/articles/)


## TimeSeries Metadata

### API Key
```{r api}
apikey <- '59cba10d8a5da536fc06b59dd0e03dee980e4c999e4f373846adcc75'
acs_names <- c("105","115","125","135","145","155","165","175","185","195","205","215","225")
```

### Metadata
```{r metadata}
time_series_mdta <- get_metadata_nhgis("time_series_tables", api_key = apikey)
time_series_mdta_acs <- time_series_mdta %>%
  filter(Reduce(`&`, lapply(acs_names, function(x) map_lgl(years, ~ x %in% .x$name))))
```

### Topics of Interest
```{r topics}
topics <- c("AV0", #Total Population
            "AV1", #Persons by Sex[2]
            "D08", #Persons by Age [2]: Children and Adults
            "B18", #Persons by Race [5*]
            "A35", #Persons of Hispanic or Latino Origin
            "CQ9", #Persons Under 18 Years in Households
            "AR5", #Total Households
            "AT5", #Persons by Nativity [2]
            "B69", #Persons 25 Years and Over by Educational Attainment [3]
            "CW0", #Total Commuters (Workers 16 Years and Over Who Did Not Work from Home)
            "C50", #Commuters by Travel Time to Work [8]
            "C98", #Aggregate Travel Time to Work for Commuters
            "BS7", #Households by Income* in Previous Year [4]
            "B79", #Median Household Income in Previous Year
            "BD5", #Per Capita Income in Previous Year
            "CL6", #Persons* below Poverty Level in Previous Year
            "C19"  #Persons* by Ratio of Income to Poverty Level in Previous Year [2]
          )
topics_notract <- c("C54", #Workers of Working Age* by Means of Transportation to Work [9]
                    "AC2" #Workers 16 Years and Over by Means of Transportation to Work [18]}
                   )
```

## API Content Download

### GIS Shape Download
```{r shape}
# Define year
shape_year <- "2022"

nhgis_shape_ext <- define_extract_nhgis(
  description = "GIS Download",
  shapefiles = c(
    #paste0("us_county_", shape_year, "_tl", shape_year),
    #paste0("us_tract_", shape_year, "_tl", shape_year),
    #paste0("us_cty_sub_", shape_year, "_tl", shape_year),
    paste0("us_place_", shape_year, "_tl", shape_year)
  )
)
nhgis_ext_submitted <- submit_extract(nhgis_shape_ext,api_key = apikey)
downloadable_extract <- wait_for_extract(nhgis_ext_submitted, api_key = apikey)
data_files <- download_extract(downloadable_extract, api_key = apikey)
```

```{r unzip}
#Note: After downloading, please move the downloaded data to the data folder, unzip its contents, and rename to exclude data request number -- example code for something to help speed up process
   # shape_url <- downloadable_extract$download_links$gis_data$url
   # shape_data <- sub(".*/", "", basename(shape_url))
   # shape_path <- file.path("data",shape_data)
   # unzip(zipfile = shape_data, exdir = shape_path)
   # file.remove(shape_data)
```

```{r boundaries}
tl2010_county <- st_read(paste0("data/nhgis_shapefile_tl",shape_year,"/nhgis_shapefile_tl",shape_year,"_us_county_",shape_year,"/US_county_",shape_year,".shp")) #%>% select(GISJOIN)

tl2010_city <- st_read(paste0("data/nhgis_shapefile_tl",shape_year,"/nhgis_shapefile_tl",shape_year,"_us_cty_sub_",shape_year,"/US_cty_sub_",shape_year,".shp")) #%>% select(GISJOIN)

tl2010_tract <- st_read(paste0("data/nhgis_shapefile_tl",shape_year,"/nhgis_shapefile_tl",shape_year,"_us_tract_",shape_year,"/US_tract_",shape_year,".shp")) #%>% select(GISJOIN)

tl2010_place <- st_read(paste0("data/nhgis_shapefile_tl",shape_year,"/nhgis_shapefile_tl",shape_year,"_us_place_",shape_year,"/US_place_",shape_year,".shp")) #%>% select(GISJOIN)
```


### Time Series Table Download
```{r acsyears}
acs_years <- c("2006-2010", "2007-2011", "2008-2012", "2009-2013", "2010-2014", 
               "2011-2015", "2012-2016", "2013-2017", "2014-2018", "2015-2019", 
               "2016-2020", "2017-2021", "2018-2022")
```

```{r rename_unzip}
rename_unzip_files <- function(download_extract, topics, i) {
  
  # determine variable names
  table_url <- download_extract$download_links$table_data$url
  table_zip <- sub(".*/", "", basename(table_url))
  table_name <-  sub("^(.*?)_.+$", "\\1", table_zip)
  name_replace <- topics[i]
  data_path <- file.path("data")
  
  # unzip downloaded data
  unzip(zipfile = table_zip, exdir = data_path)
  
  # all items to rename
  all_items <- c(list.files(data_path, full.names = TRUE, recursive = TRUE),
                 list.dirs(data_path, full.names = TRUE, recursive = TRUE))

  #iIterate over each item
  for (item in all_items) {
    base_name <- basename(item)
    new_name <- gsub(table_name, name_replace, base_name)
    new_name <- gsub("_csv", "", new_name)
    new_path <- file.path(dirname(item), new_name)
    file.rename(item, new_path)
  }

  #remove original zip file
  file.remove(table_zip)
}
```

```{r download}
download_data <- function(t,g,a){
  
  for (i in seq_along(t)) {
    # format request
    nhgis_ext2 <- define_extract_nhgis(
    description = "Time Series Download",
    time_series_tables = tst_spec(
      t[i], 
      geog_levels = g,
      years = a
      )
    )
  
    # submit and download requested data
    nhgis_ext_submitted2 <- submit_extract(nhgis_ext2,api_key = apikey)
    downloadable_extract2 <- wait_for_extract(nhgis_ext_submitted2, api_key = apikey)
    data_files2 <- download_extract(downloadable_extract2, api_key = apikey)
  
    rename_unzip_files(downloadable_extract2, t, i)
  }
}
```

```{r}
tops = c("AC2") #topics
geogs = c("place") 

download_data(tops,geogs, acs_years)
```

## Join Shape to Time Series







## Still to Do
- get all the time series data that exists at the places level
- write api to get OTHER data.. maybe (other than time series)
- join to 2022 shapefile -- unless its the one standardized to 2010, then use 2010
- upload to AGOL
- meet with Bert for next steps


## Bert Notes
- Focus more time on 1 or 2 test cases than downloaded a bunch of stuff (SOV vehicle commutes, or telecommuting percentage)
- places for time series data with acs 5 year
- counties of interest tooele, morgan, summit, wasatch, davis, weber, salt lake, utah, box elder


## Chad Notes
- It might be a good idea to "double check" and review the 1_Inputs/2_SEData/_ControLTotals/_Source - ControlTotal_SE - 2022-08-31.xlsx (Sheet Process_SE_Historic) to make sure the values match up with NHGIS data
- Telecommuting is not a variable within ACS or census questions. Instead they use Work at Home, and in some cases they use different self-employment variables. Telecommute (according to the process of the TDM) is calculated as the Total Work at Home value minus the self-employment totals. 
  - tables to use: MEANS OF TRANSPORTATION TO WORK BY CLASS OF WORKER, MEANS OF TRANSPORTATION TO WORK BY INDUSTRY
  




